---
title: Count distinct occurrences of a variable by another variable
author: Mauro Lepore and Monique Picon
date: '2018-07-30'
slug: count-distinct-occurrences-of-a-variable-by-another-variable
categories: []
tags:
  - R
  - dplyr
  - tidyverse
  - count
description: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

We want to count the number of species (`sp`) in each `site`. Here we'll use a dataset that is so small that we can count by eye. But in real life you'll need to approach this problem programatically. Here is how.

* Setup

```{r}
library(tidyverse)
```

* Data: Site A has only one species; site B has three species.

```{r}
dataset <- tibble::tribble(
  ~site,  ~sp, ~other_vars,
    "A", "sp1",          1,
    "A", "sp1",          1,
    "A", "sp1",          2,
    "B", "sp1",          3,
    "B", "sp2",          4,
    "B", "sp3",          5
)
dataset
```

* Option 1: Expressive way. Group the data by `site` and count the unique occurrences of `sp`.

```{r}
dataset %>% 
  group_by(site) %>% 
  summarise(n = n_distinct(sp))
```

* Option 2: A bit more cryptic way: Select the relevant columns; get the unique combinations of values; count the number of rows by `site`.

```{r}
dataset %>% 
  select(site, sp) %>% 
  unique() %>% 
  count(site)
```

Warning: This is wrong. If you don't understand why, then use the more expressive approach (option 1 above).

```{r}
dataset %>% 
  select(site, sp) %>% 
  unique() %>% 
  # This is wrong! You should count by site -- not species.
  count(sp)
```

