---
title: Gathering and spreading
author: Mauro Lepore
date: '2018-07-28'
slug: gathering-and-spreading
categories: []
tags:
  - R
  - tidyr
description: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

```{r}
library(tidyverse)
```

# Gathering

ForestGEO datasets encode the status of each stem in the values of the column `status`.

```{r}
select(fgeo.data::luquillo_stem5_random, status)
```
In raw data, however, the values of `status` are sometimes spread across multiple columns. For example:

```{r}
wide <- tibble::tribble(
  ~treeID, ~stemID,    ~gx,    ~gy, ~dbh,  ~A,  ~G, ~P,
     104L,    143L,  10.31, 245.36,  193, "A",  NA, NA,
     119L,    158L, 182.89, 410.15, 40.4, "A",  NA, NA,
     180L,    222L, 164.61,  409.5,   NA,  NA, "G", NA,
     180L,    223L, 164.61,  409.5,   NA,  NA, "G", NA,
     180L,    224L, 164.61,  409.5,   NA,  NA, "G", NA,
     180L,    225L, 164.61,  409.5,   45, "A",  NA, NA,
     602L,    736L, 148.96, 414.44,   33, "A",  NA, NA,
     631L,    775L,   38.3,  245.3,  140, "A",  NA, NA,
     647L,    793L, 143.24, 410.91,  246, "A",  NA, NA,
    1086L,   1339L,  68.89, 253.04,  176, "A",  NA, NA
)

wide
```

We can gather the multiple `status` columns with `tidyr::gather()`. See `?gather()` for a definition of each argument.

```{r}
long <- gather(wide, key = "status", "value", A, G, P, na.rm = TRUE)
long

# In this case the column `value` is useless
select(long ,-value)
```

The raw data may also look like this.

```{r}
wide2 <- tibble::tribble(
  ~treeID, ~stemID,    ~gx,    ~gy, ~dbh, ~A, ~G, ~P,
     104L,    143L,  10.31, 245.36,  193,  1,  0,  0,
     119L,    158L, 182.89, 410.15, 40.4,  1,  0,  0,
     180L,    222L, 164.61,  409.5,   NA,  0,  1,  0,
     180L,    223L, 164.61,  409.5,   NA,  0,  1,  0,
     180L,    224L, 164.61,  409.5,   NA,  0,  1,  0,
     180L,    225L, 164.61,  409.5,   45,  1,  0,  0,
     602L,    736L, 148.96, 414.44,   33,  1,  0,  0,
     631L,    775L,   38.3,  245.3,  140,  1,  0,  0,
     647L,    793L, 143.24, 410.91,  246,  1,  0,  0,
    1086L,   1339L,  68.89, 253.04,  176,  1,  0,  0
)

wide2
```

```{r}
long2 <- gather(wide2, key = "status", value = "value", A, G, P)
# Sorting to better understand the point that follows
arrange(long2, treeID, stemID)
```

We need to remove duplicates. For each stem the number of rows should be unique but instead we have 3 rows -- one for each of the columns we gathered. Above we could fix this by using `na.rm`. Now we have no missing values but we can get the same result by excluding the rows where `value` is zero.

```{r}
filter(long2, value != 0)

# We no longer need `value` so we can remove it

single_rows <- filter(long2, value != 0)
single_rows <- select(single_rows, -value)
single_rows
```

# Spreading

The opposite of gathering is spreading. This is not a common operation in ForestGEO data but I'll it for completeness.

```{r}
spread(long2, key = "status", value = "value")
```

