---
title: Calculate and plot proportions in multiple facets of a dataset
author: Mauro Lepore and Monique Picon
date: '2018-08-03'
slug: calculate-and-plot-proportions-in-multiple-facets-of-a-dataset
categories: []
tags:
  - ggplot2
  - R
description: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)

knitr::opts_chunk$set(
  fig.path = "public/post/2018-08-03-calculate-and-plot-proportions-in-multiple-facets-of-a-dataset_files/",
  echo = TRUE,
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  # Figures (http://r4ds.had.co.nz/)
  out.width = "95%",
  fig.align = "center",
  fig.width = 6,
  fig.asp = 0.618,
  fig.show = "hold"
)
```

How would you plot the proportion of observations by a categorical variable, for multiple facets of your data? Here we do it with the tidyverse.

```{r}
library(tidyverse)
```

We use a (private) dataset on the damage recorded on dead and alive trees at three different sites.
 
```{r, echo=FALSE, message=FALSE, warning=FALSE}
dmg <- read_csv(here::here("content/post/data/private/damage10.csv"))
dmg
```

### Clean data

(This section is mainly cosmetic -- not the main point of this post.)

We recode the values of three variables to make them more informative. 

```{r, fig.align="default", out.width="50%", fig.widh=(6 * 0.5 / 0.7)}
dmg <- dmg %>% 
  mutate(
    site = case_when(
      site == "EV1" ~ "M1",
      site == "SB1" ~ "M2",
      site == "SB3" ~ "OG"
    )
  ) %>% 
  mutate(
    b_u = case_when(
      b_u == "B" ~ "broken",
      b_u == "U" ~ "uprooted",
      b_u == "B/U" ~ "both",
      is.na(b_u) ~ "none"
    )
  ) %>% 
  mutate(
    status = case_when(
      status == "A" ~ "Alive",
      status == "D" ~ "Dead"
    )
  )
```

We also rename a variable, make it a factor, and reorder its levels, so that they appear in the plot in the order that best tells the story we want to tell with this dataset.

```{r}
dmg <- dmg %>% 
  rename(stem_damage = b_u) %>%
  mutate(stem_damage = lvls_reorder(factor(stem_damage), c(4, 2, 1, 3)))
dmg
```

Now we compute on each site separately. We create a helper function to calculate the proportion of stem damage in each status category. Our helper does the computation for one site.

```{r}
# Calculate the proportion of each damage category in a single site
proportion_each_site <- function(dmg_site) {
  dmg_site %>% 
    select(status, stem_damage) %>% 
    count(status, stem_damage) %>% 
    # The column `n` results from `count()`
    mutate(total = sum(n, na.rm = TRUE), proportion = n / total)
}
```

Now we apply the helper function to each site with `map()`. We use `nest()` to work within a dataframe.

```{r}
dmg_prop <- dmg %>% 
  nest(-site) %>% 
  mutate(data = map(data, ~proportion_each_site(.x))) %>% 
  unnest()
dmg_prop
```

Finally we plot each facet of our data in a different panel.

```{r}
dmg_prop %>% 
  ggplot(aes(status, proportion)) +
    geom_col(aes(fill = stem_damage)) +
    facet_wrap(vars(site))
```

Note that the values within each panel add to 1.

### What did NOT work

We first calculated the proportion for each site directly within the ggplot with `stat()`, hoping that the calculation would be applied to each facet of the data independently. But this didn't work: The proportion was calculated not for each facet independently but for the entire dataset.

```{r}
ggplot(dmg, aes(status)) +
  facet_wrap(vars(site)) +
  geom_bar(aes(fill = stem_damage, y = stat(count / sum(count)))) +
  labs(y = "proportion")
```

If you know how to adapt this approach to plot what we want please let us know.
